Evolution of pre-copulatory and post-copulatory inbreeding behaviour

### 2016-10-21 ms_pre_post.Rmd
Full edits for new draft; still need to redo figure 2

We track interactions among discrete individuals in a small focal population that persists over multiple generations (key individual traits, parameter values, and variables are described in Table 1). Individuals within this population are either female or male, and can be related to each other if they share a common ancestor. We record the coefficient of kinship ($k$), defined as the probability that a randomly selected homologous allele from each individual is identical-by-descent [@Lynch1998], between all pairs of individuals in each generation using the population pedigree [see @Boyce1983; @Duthie]. At the start of each generation, each female pays up to three independent costs of (1) tendency to engage in polyandry ($c_{P}$), (2) pre-copulatory inbreeding strategy (i.e., cost of mate choice; $c_{M}$), and (3) post-copulatory inbreeding strategy (i.e., cost of fertilisation bias; $c_{F}$). The number of males that a focal female mates with ($N_{males}$) is determined by the effects of 20 alleles ($P_{a}$) across 10 diploid loci that combine additively to determine her genotype ($P_{g}$), and ultimately her phenotype for tendency to engage in polandry ($P_{p}$). A focal female's choice of mate(s) is affected by 20 alleles ($M_{a}$) across 10 diploid loci that combine additively to affect her genotype ($M_{g}$) and phenotype ($M_{p}$) for pre-copulatory inbreeding avoidance or preference, and kinship between the focal female and her potential mates. Similarly, after mating has occurred, post-copulatory fertilisation of a focal female's $n$ total offspring is affected by 20 alleles ($F_{a}$) across 10 diploid loci that combine additively to affect the genotype ($F_{g}$) and phenotype ($F_{p}$) of the degree to which fertilisation success is affected by kinship. After fertilisation, the viability of a focal female's $n$ offspring is decreased by inbreeding depression ($\beta$), and surviving offspring become adults. After inbreeding depression in offspring viability, $\rho$ adult male immigrants arrive in the population, then density effects reduce the entire population down to separate female ($K_{f}$) and male ($K_{m}$) carrying capacities. Over generations, we record mean values of alleles underlying polyandry, pre-copulatory inbreeding strategy, and post-copulatory inbreeding strategy.

@@

Each male within the population is a potential mate for any female to choose. We therefore assume that there are no opportunity costs associated with male mating such that mating with one female would reduce a male's ability to mate with and sire the offspring of any other female [e.g., @Waser1986]. While all males are therefore available as potential mates, we assume that a female mates with her total allotment of $N_{males}$ without replacement, meaning that $N_{males}$ models the total number of distinct males that a female chooses to mate with rather than the number of times that she mates. In the unlikely event that a female's $N_{males}$ exceeds the total number of males in the population, then she mates with all of the males in the populaton. Most often, $N_{males}$ will be a small subset of the available males in the population [@Duthie], so a focal female will be able to preferentially mate with males that are close kin (high $k$), distant kin (low $k$), or non-kin ($k=0$) through the expression of her pre-copulatory inbreeding strategy phenotype ($M_{p}$). Like polyandry allele values, pre-copulatory inbreeding strategy allele values ($M_{a}$) can take any real number. However, unlike polyandry, the sum of all of a female's 20 $M_{a}$ alleles equals both her pre-copulatory inbreeding strategy genotype ($M_{g}$) and her pre-copulatory inbreeding strategy phenotype ($M_{p}$). Negative or positive values of $M_{p}$ cause a female to avoid or prefer mating with kin, respectively. Values of $M_{p}=0$ cause females to mate randomly with respect to kinship.

@@

Alleles of a focal female's offspring mutate with a probability of $\mu=0.001$, with mutation being an independent event that does or does not occur for each locus. In general, loss of function mutations might be expected to be more frequent than mutations that increase the functionality of adaptations that rely on complex physiologies or behaviours. If tendency for polyandry, inbreeding mating strategies, or inbreeding fertilisation strategies rely on such complex physiologies or behaviours, then we might expect the direction of mutation to be biased (e.g., towards monandry or random mating). However, because we are interested specifically in how selection affects reproductive strategy given varying costs and inbreeding depression, we model mutation as unbiased so that selection can be isolated and inferred clearly from mean allele values.

Hence, when a mutation occurs at a locus, a new value is sampled from a random normal distribution with a mean of zero and a standard devation of mutation effect size $\mu_{SD}$, $\mathcal{N}(0, \mu_{SD})$. This new value is then added to the original value of the allele that mutates [@Kimura1965; @Lande1976; @Bocedi2014; @Duthie]. The standard deviation of the mutation effect size ($\mu_{SD}$) is set to $\sqrt{1/20}$, which scales the variance of mutation effect size to the total number of alleles affecting $P_{p}$, $M_{p}$, and $F_{p}$ phenotypes [see @Kimura1965; @Lande1976; @Bocedi2014].

@@

Selection for polyandry is driven by pre-copulatory inbreeding avoidance
----------------------------------------------

For low costs of tendency for polyandry, $c_{P} < 0.005$, selection caused expected tendency for polyandry allele values ($P_{a}$) to be positive only when post-copulatory inbreeding avoidance evolved (Figure 1A-D), and for all $c_{P}$, $P_{a}$ values were higher when post-copulatory inbreeding avoidance evolved than when it could not evolve (Figure 1). When post-copulatory inbreeding avoidance could not evolve, $P_{a}$ values were negative even in the absence of a direct cost ($c_{P}=0$; Figure 1A) because females that expressed polyandry were more likely to increase their inbreeding due to sampling males without replacement [@Duthie]. Strong post-copulatory inbreeding avoidance consistently evolved in all simulations, even when $P_{a}$ values were expected to be slightly negative (Figure 1F,H), because of the threshold nature of the polyandry phenotype; Figure 2 shows how phenotypic polyandry was common even when mean polyandry allele values within a population were expected to be negative. In all simulations, pre-copulatory inbreeding avoidance also evolved ($M_{a}<0$), meaning that both pre- and post-copulatory inbreeding avoidance was predicted, and evolution of adaptive polyandry occurred given sufficiently low costs of tendency for polyandry.

Selection greatly weakened for a relatively more costly inbreeding strategy
----------------------------------------------

For simulations in which either pre- or post-copulatory inbreeding strategy allele values were fixed for random choice with respect to kinship ($M_{a}=0$ or $F_{a}=0$, respectively), the alternative costly inbreeding strategy evolved to inbreeding avoidance ($F_{a}<0$ or $M_{a}<0$; Figure 3). However, when the alternative inbreeding strategy could evolve and was not costly, the focal costly inbreeding strategy initially evolved towards inbreeding avoidance, but then evolved toward random choice given increasingly negative allele values of the alternative strategy over generations (Figure 3B,D). For example, given fixed $F_{a}=0$, grand mean $M_{a}$ values were $-0.868$ in the last $10000$ generations (Figure 3A), meaning that an outbred focal female was ca $5.34$ times less likely to choose a full-sibling and ca $2.10$ times less likely to choose a first cousin than a non-relative given a strong cost of pre-copulatory inbreeding avoidance, $c_{M}=0.02$. In contrast, grand mean $M_{a}$ allele values were $-0.288$ when post-copulatory inbreeding avoidance also evolved. More strongly, given fixed $M_{a}=0$, grand mean $F_{a}$ values were $-1.092$ in the last $10000$ generations (Figure 3C), meaning that an outbred focal female was ca $6.46$ times less likely to choose a full-sibling and ca $2.37$ times less likely to choose a first cousin than a non-relative given a strong cost of pre-copulatory inbreeding avoidance, $c_{F}=0.02$. However, grand-mean $F_{a}$ values were $-0.070$ when pre-copulatory inbreeding strategy evolved, and polyandry allele values were negative. Overall, selection for a costly inbreeding strategy that was adaptive in isolation was greatly weakened given evolution of a less costly strategy.

Existing adaptation preclude post- but not pre-copulatory inbreeding avoidance
----------------------------------------------

@@

We modelled evolution of three separate traits affecting polyandry and inbreeding avoidance, including tendency to engage in polyandry, pre-copulatory inbreeding strategy, and post-copulatory inbreeding strategy. Using *in silico* experiments, we tested three specific hypothesis to predict evolution of polyandry, and pre- and post-copulatory inbreeding strategy. We found that (1) selection for polyandry was greatly strengthened given concurrent evolution of post-copulatory inbreeding avoidance, (2) selection for costly pre- or post-copulatory inbreeding avoidance was negligible given evolution of a less costly strategy of inbreeding avoidance, and (3) pre-existence of adaptive pre-copulatory inbreeding avoidance precluded evolution of polyandry and post-copulatory inbreeding avoidance, but not vice versa. More generally, we provide a novel framework for approaching empirical studies of inbreeding avoidance, predicting evolution of polyandry and inbreeding avoidance through pre- and post-copulatory mechanisms.

@@

Specific mechanisms of post-copulatory inbreeding avoidance are difficult to identify, but recent experimental studies have shown how females can successfully bias fertilisation after mating has occurred in both vertebrates [e.g., @Gasparini2011] and invertebrates [e.g., @Tuni2013]. For example, @Gasparini2011 artificially inseminated female guppies (*Poecilia reticulata*) with equal quantities of sperm from full-siblings and unrelated males, finding that fertilisation success was higher in non-relatives because the velocity of full siblings' sperm was reduced by females' ovarian fluids. Biased fertilisation toward non-relatives might also be accomplished through selective storage of sperm. For example, in black field crickets (*Teleogryllus commodus*), females attempt to remove the spermataphores of unwanted male mates after copulation, while male mates guard females to prevent spermatophore removal [@Bussiere2006]. Even if females are unable to remove spermatophores of unwanted mates, @Tuni2013 found that female *T. commodus* are capable of controlling the transfer of sperm to their spermatheca, most likely through physical control of the spermathecal duct, and can reduce sperm storage despite being guarded by a related male. Our results suggest that evolution of such post-copulatory inbreeding avoidance might increase selection for polyandry, especially if pre-copulatory inbreeding avoidance is costly (Figure 1).

In our model, a cost free inbreeding strategy precluded a more costly complementary inbreeding strategy from persisting in a focal population (Figure 2), meaning that long-term persistence of both pre- and post-copulatory inbreeding avoidance might not be expected in populations given cost asymmetry. Direct costs therefore strongly affected the evolution of the costly phenotype(s), meaning that quantifying costs is necessary for predicting evolution of phenotypes associated with inbreeding strategies. Quantifying direct costs of phenotypes associated with polyandry and mate choice is empirically challenging [@Pomiankowski1987; @Kokko2003; @Reid2015]. @Pomiankowski1987 categorised four types of costs that are relevant to mate choice, which we suggest are also relevant costs for polyandry, and include costs of elevated risks associated with (1) predation or (2) disease transmission, and costs incurred through (3) loss of time or (4) depletion of energy. In our model, we interpreted a cost of tendency for polyandy as an elevated risk of predation as might be incurred while searching for or courting mates [e.g., @Rowe1988; @Rowe1994], a cost of pre-copulatory inbreeding strategy as a lost of time [i.e., risk of not finding a mate in time due to choosiness; e.g., @Kokko2013], and a cost of post-copulatory inbreeding strategy as depletion of energy. While these costs are biologically realistic, different costs will be relevant for different populations and be realised in different ways. For example, increased risk of disease transmission has been observed for highly polyandrous females [@Roberts2015a], but such a cost would more realistically be applied to the number of realised mates a female has rather than her tendency for engaging in polyandry. Future models will therefore need to carefully consider the magnitudes and realisation of direct costs to accurately predict evolution of inbreeding strategies in populations. <!--- add Tuni citation for depletion of energy? --->

@@

Our model assumed that allele values underlying tendency for polyandry, pre-copulatory inbreeding avoidance, and post-copulatory inbreeding avoidance were all initialised at zero (causing monandry, and random mating and fertilisation with respect to kinship), and phenotypes evolved simultaneously after initialisation. However, these initial conditions are likely to be atypical, as evolution of pre-copulatory and post-copulatory inbreeding avoidance from ancestral random mating and fertilisation is unlikely to occur simultaneously. Rather, initial evolution of one inbreeding strategy (e.g., pre-copulatory inbreeding avoidance) is more likely to occur in the absence of its complementary strategy (e.g., post-copulatory inbreeding avoidance) if the complementary strategy is evolutionarily constrained. Alternatively, initial evolution of one inbreeding strategy might occur in the context of an already established adaptive strategy for avoiding inbreeding. We showed that selection for post-copulatory but not pre-copulatory inbreeding avoidance was strengthened if its complementary strategy was fixed for random choice with respect to kinship, and that selection for post-copulatory but not pre-copulatory inbreeding avoidance was precluded if its complementary strategy was fixed for adaptive inbreeding avoidance.

@@

\vspace{80 mm}

>>>>

We model evolution of polyandry and inbreeding avoidance through pre- and post-copulatory mechanisms by tracking interactions among discrete individuals in a small focal population that persists over multiple generations (key individual traits, parameter values, and variables are described in Table 1). Individuals within this population are either female or male, and can be related to each other if they share a common ancestor. Each individual has 10 diploid loci and therefore 20 alleles underlying each trait: polyandry ($P_{a}$), pre-copulatory inbreeding avoidance ($M_{a}$), and post-copulatory inbreeding avoidance ($F_{a}$). Alleles combine additively to affect genotypes ($P_{g}$, $M_{g}$, and $F_{g}$) and phenotypes ($P_{p}$, $M_{p}$, and $F_{p}$) for each trait. Over generations, we record mean values of alleles underlying polyandry, pre-copulatory inbreeding strategy, and post-copulatory inbreeding strategy. One generation proceeds with females paying costs, expressing polyandry, mating, and fertilising offspring; offspring alleles then mutate and offspring express inbreeding depression; finally, immigrants arrive in the populatoin and density regulation limits population growth.

@@

Each male within the population is a potential mate for any female to choose. We therefore assume that there are no opportunity costs associated with male mating such that mating with one female would reduce a male's ability to mate with and sire the offspring of any other female [e.g., @Waser1986]. While all males are therefore available as potential mates, we assume that a female mates with her total allotment of $N_{males}$ without replacement, meaning that $N_{males}$ models the total number of distinct males that a female chooses to mate with rather than the number of times that she mates. In the unlikely event that a female's $N_{males}$ exceeds the total number of males in the population, then she mates with all of the males in the populaton. Most often, $N_{males}$ will be a small subset of the available males in the population [@Duthie], so the male(s) a focal female chooses will reflect her expression of pre-copulatory inbreeding strategy phenotype ($M_{p}$). Like polyandry allele values, pre-copulatory inbreeding strategy allele values ($M_{a}$) can take any real number. However, unlike polyandry, the sum of all of a female's 20 $M_{a}$ alleles equals both her pre-copulatory inbreeding strategy genotype ($M_{g}$) and her pre-copulatory inbreeding strategy phenotype ($M_{p}$). Negative or positive values of $M_{p}$ cause a female to avoid or prefer mating with kin, respectively. Values of $M_{p}=0$ cause females to mate randomly with respect to kinship.

@@

Alleles of a focal female's offspring mutate with a probability of $\mu=0.001$, with mutation being an independent event that does or does not occur for each locus. When a mutation occurs at a locus, a new value is sampled from a random normal distribution with a mean of zero and a standard devation of mutation effect size $\mu_{SD}$, $\mathcal{N}(0, \mu_{SD})$. This new value is then added to the original value of the allele that mutates [@Kimura1965; @Lande1976; @Bocedi2014; @Duthie]. The standard deviation of the mutation effect size ($\mu_{SD}$) is set to $\sqrt{1/20}$, which scales the variance of mutation effect size to the total number of alleles affecting $P_{p}$, $M_{p}$, and $F_{p}$ phenotypes [see @Kimura1965; @Lande1976; @Bocedi2014].

@@

Will evolution of post-copulatory inbreeding avoidance stratengthen selection for polyandry?
----------------------------------------------

For low costs of polyandry, $c_{P} < 0.005$, selection caused expected polyandry allele values ($P_{a}$) to be positive only when post-copulatory inbreeding avoidance evolved (Figure 1A-D). For all $c_{P}$, $P_{a}$ values were higher when post-copulatory inbreeding avoidance evolved than when it could not evolve (Figure 1). When post-copulatory inbreeding avoidance could not evolve, $P_{a}$ values were negative even in the absence of a direct cost ($c_{P}=0$; Figure 1A) because females that expressed polyandry were more likely to increase their inbreeding due to sampling males without replacement [@Duthie]. Strong post-copulatory inbreeding avoidance consistently evolved in all simulations, even when $P_{a}$ values were expected to be slightly negative (Figure 1F,H), because of the threshold nature of the polyandry phenotype; Figure 2 shows how phenotypic polyandry was common even when mean polyandry allele values within a population were expected to be negative. In all simulations, pre-copulatory inbreeding avoidance also evolved ($M_{a}<0$), meaning that both pre- and post-copulatory inbreeding avoidance was predicted, and evolution of adaptive polyandry occurred given sufficiently low costs of polyandry.

Will selection for a costly inbreeding strategy be negligible given a non-costly alternative strategy?
----------------------------------------------

For simulations in which either pre- or post-copulatory inbreeding strategy allele values were fixed for random choice with respect to kinship ($M_{a}=0$ or $F_{a}=0$, respectively), the alternative costly inbreeding strategy evolved to inbreeding avoidance ($F_{a}<0$ or $M_{a}<0$; Figure 3). However, when the alternative inbreeding strategy could evolve and was not costly, the focal costly inbreeding strategy initially evolved towards inbreeding avoidance, but then evolved toward random choice with respect to kinship given increasingly negative allele values of the alternative strategy over generations (Figure 3B,D). For example, given fixed $F_{a}=0$, grand mean $M_{a}$ values were $-0.868$ in the last $10000$ generations (Figure 3A), meaning that an outbred focal female was ca $5.34$ times less likely to choose a full-sibling and ca $2.10$ times less likely to choose a first cousin than a non-relative given a strong cost of pre-copulatory inbreeding avoidance, $c_{M}=0.02$. In contrast, grand mean $M_{a}$ allele values were $-0.288$ when post-copulatory inbreeding avoidance also evolved. More strongly, given fixed $M_{a}=0$, grand mean $F_{a}$ values were $-1.092$ in the last $10000$ generations (Figure 3C), meaning that an outbred focal female was ca $6.46$ times less likely to choose a full-sibling and ca $2.37$ times less likely to choose a first cousin than a non-relative given a strong cost of pre-copulatory inbreeding avoidance, $c_{F}=0.02$. However, grand-mean $F_{a}$ values were $-0.070$ when pre-copulatory inbreeding strategy evolved, and polyandry allele values were negative. Overall, selection for a costly inbreeding strategy that was adaptive in isolation was greatly weakened given evolution of a less costly strategy.

Will selection for a costly inbreeding strategy be negligible given existence of an already adaptive alternative strategy?
----------------------------------------------

@@

We have provided a novel framework for approaching empirical studies of inbreeding avoidance, predicting evolution of polyandry and inbreeding avoidance through pre- and post-copulatory mechanisms. Accordingly, we modelled evolution of three separate traits affecting polyandry and inbreeding avoidance, including tendency to engage in polyandry, pre-copulatory inbreeding strategy, and post-copulatory inbreeding strategy. Using *in silico* experiments, we tested three specific hypothesis to predict evolution of polyandry, and pre- and post-copulatory inbreeding strategy. We found that (1) selection for polyandry was greatly strengthened given concurrent evolution of post-copulatory inbreeding avoidance, (2) selection for costly pre- or post-copulatory inbreeding avoidance was negligible given evolution of a less costly strategy of inbreeding avoidance, and (3) pre-existence of adaptive pre-copulatory inbreeding avoidance precluded evolution of polyandry and post-copulatory inbreeding avoidance, but not vice versa.

@@

Specific mechanisms of post-copulatory inbreeding avoidance are difficult to identify, but recent experimental studies have shown how females can successfully bias fertilisation after mating has occurred in both vertebrates [e.g., @Gasparini2011] and invertebrates [e.g., @Tuni2013]. For example, @Gasparini2011 artificially inseminated female guppies (*Poecilia reticulata*) with equal quantities of sperm from full-siblings and unrelated males, finding that fertilisation success was higher in non-relatives because the velocity of full siblings' sperm was reduced by females' ovarian fluids. Biased fertilisation toward non-relatives might also be accomplished through selective storage of sperm. For example, in black field crickets (*Teleogryllus commodus*), females attempt to remove the spermataphores of unwanted male mates after copulation, while male mates guard females to prevent spermatophore removal [@Bussiere2006]. Even if females are unable to remove spermatophores of unwanted mates, @Tuni2013 found that female *T. commodus* are capable of controlling the transfer of sperm to their spermatheca, most likely through physical control of the spermathecal duct, and can reduce sperm storage despite being guarded by a related male. Our results suggest that evolution of such post-copulatory inbreeding avoidance might increase selection for further polyandry, especially if pre-copulatory inbreeding avoidance is costly (Figure 1).

In our model, a cost free inbreeding strategy precluded a more costly complementary inbreeding strategy from persisting in a focal population (Figure 2), meaning that long-term persistence of both pre- and post-copulatory inbreeding avoidance might not be expected in populations given cost asymmetry. Direct costs therefore strongly affected the evolution of the costly phenotype(s), meaning that quantifying costs is necessary for predicting evolution of phenotypes associated with inbreeding strategies. Quantifying direct costs of phenotypes associated with polyandry and mate choice is empirically challenging [@Pomiankowski1987; @Kokko2003; @Reid2015]. @Pomiankowski1987 categorised four types of costs that are relevant to mate choice, which we suggest are also relevant costs for polyandry, and include costs of elevated risks associated with (1) predation or (2) disease transmission, and costs incurred through (3) loss of time or (4) depletion of energy. In our model, we interpreted a cost of polyandy as an elevated risk of predation as might be incurred while searching for or courting mates [e.g., @Rowe1988; @Rowe1994], a cost of pre-copulatory inbreeding strategy as a lost of time [i.e., risk of not finding a mate in time due to choosiness; e.g., @Kokko2013], and a cost of post-copulatory inbreeding strategy as depletion of energy. While these costs are biologically realistic, different costs will be relevant for different populations and be realised in different ways. For example, increased risk of disease transmission has been observed for highly polyandrous females [@Roberts2015a], but such a cost would more realistically be applied to the number of realised mates a female has rather than her tendency for engaging in polyandry. Future models will therefore need to carefully consider the magnitudes and realisation of direct costs to accurately predict evolution of inbreeding strategies in populations. <!--- add Tuni citation for depletion of energy? --->

@@

Our model assumed that allele values underlying polyandry, pre-copulatory inbreeding avoidance, and post-copulatory inbreeding avoidance were all initialised at zero (causing monandry, and random mating and fertilisation with respect to kinship), and phenotypes evolved simultaneously after initialisation. However, these initial conditions are likely to be atypical, as evolution of pre-copulatory and post-copulatory inbreeding avoidance from ancestral random mating and fertilisation is unlikely to occur simultaneously. Rather, initial evolution of one inbreeding strategy (e.g., pre-copulatory inbreeding avoidance) is more likely to occur in the absence of its complementary strategy (e.g., post-copulatory inbreeding avoidance) if the complementary strategy is evolutionarily constrained. Alternatively, initial evolution of one inbreeding strategy might occur in the context of an already established adaptive strategy for avoiding inbreeding. We showed that selection for post-copulatory but not pre-copulatory inbreeding avoidance was strengthened if its complementary strategy was fixed for random choice with respect to kinship, and that selection for post-copulatory but not pre-copulatory inbreeding avoidance was precluded if its complementary strategy was fixed for adaptive inbreeding avoidance.

@@

\vspace{140 mm}
